The yaq Project: \\ A Standard for Scientific Instrumentation

### 2022-10-21 paper.tex
morris comments

%  1)  latex  paper
%  2)  bibtex paper
%  3)  latex  paper
%  4)  latex  paper
%
\documentclass[aip, amsmath, amssymb, reprint,]{revtex4-1}

@@

Modern instrumentation development often involves incorporation of many dissimilar hardware peripherals into a single unified instrument.
Increasing hardware availability has brought greater instrument complexity to small research groups.
The flexibility promised to instrument makers by large hardware availability stretches the capability of traditional monolithic orchestration software.
In many cases, a lack of software flexibility leads creative researchers to feel frustrated and unable to perform the experiments thy envision.
Highly modular hardware components enable extreme experimental flexibility.
Herein, we describe \yaq{}, a software project defining a new standard way of communicating with diverse hardware peripherals.
\yaq{} encourages a highly modular approach to experimental software development, which is well suited to address the experimental flexibility needs of such instruments.
\yaq{} is architected to overcome hardware communication barriers that are insurmountable with typical experimental software.
A large number of hardware peripherals are already supported with tooling available to create more.
The \yaq{} standard enables collaboration between multiple research groups, increasing code quality while lowering development effort.

@@

Instrumentation development is a key part of the scientific enterprise.
Novel instruments are typically constructed of many individual components, purchased and home-built.
Orchestration software must communicate with each hardware component in the course of a scientific experiment.
This can involve utilization of many interfaces: NI DAQmx\cite{nidaqmx}, SCPI\cite{scpi}, ModBus\cite{modbus}, PICam\cite{picam}, Thorlabs APT\cite{thorlabs_apt}, among many others.
The challenge of integrating all of these is a frustrating piece of the modern instrument development process.
Weeks can be spent just integrating one new component into an existing project.
Code reusability is typically poor, and technical debt grows quickly in academic and educational contexts.
Scientists may struggle to rapidly innovate on their experimental design when each hardware addition requires major software development.

Some large user-facilities have addressed interface complexity via the adoption unified standards, such as EPICS \cite{DalesioLR1991a} or TANGO \cite{AGotz1999TANGOA}.
The unified standards define a network interface for any hardware component.
Orchestration software can target this unified standards for reading and writing hardware state.
Small background services are written to translate the myriad component interfaces into the standard EPICS IOCs and TANGO Devices.
These programs are performant, but require expert management to set up and provide descriptions via a separate server program.
These projects are impressive at facility scale, but in our experience these do not scale well to single-investigator lab environments.

@@

Here, we discuss the design of the \yaq{} standard in the context of four challenges facing instrument designers.
First, we discuss how particularly challenging hardware interfaces can become seemingly insurmountable barriers to software control.
Next, we discuss how inflexible orchestration software can limit experimental creativity.
Next, we focus on challenges that arise when enhancing or modifying existing instruments with new hardware.
Finally, we discuss the heavy software maintenance burden that many instrument designers face.
In each case, we will highlight how the \yaq{} project is designed to alleviate that challenge.
Several case studies provide a view into the flexible ways that \yaq{} can be applied to perform different scientific experiments.
This document provides an overview of the concepts and motivations behind \yaq{}, but does not serve as a formal specification\cite{yaq_fyi}.

@@

Hardware-software interface design presents several technical challenges.
In this section we describe the architecture of the \yaq{} framework in light of three major barriers that we have encountered in scientific instrumentation development.
First barrier: Multiple interfaces are used to communicate with each component of the system.
A fully automated system must be able to use all of these interfaces, a daunting task for scientists who do not specialize in software development.
Second barrier: Certain specialty hardware have inconvenient interface requirements.
A camera will only work with an obsolete interface card and drivers for Windows XP.
A data acquisition manufacturer provides an Application Programmer Interface (API) that only works in Python 3.7.
A graduate student wishes to drive several stepper motors using a Raspberry Pi.
Third barrier: Some hardware interfaces are blocking.
A graphical user interface stalls while waiting for a camera to collect data.
Custom orchestration software needs to be closed before the manufacturers configuration software can be used.
A graduate student finds themself needing to master advanced concepts in concurrency in order to orchestrate many motors performantly.

@@

Figure \ref{fig:network} diagrams the \yaq{} architecture.
Here, we show three different computers connected via an ethernet network.
The top and bottom computer are connected to monitors for interactive use while the middle computer is only accessible via the network.
This might represent a complex scientific instrument involving several operator terminals as well as embedded computers.
At the top, a single computer is connected to four hardware peripherals through RS232, TTL, USB, and PCI as indicated by the colored lines.
That same computer is running four separate programs, one for each peripheral.
These small, targeted, programs are managed by the operating system and run in the background.
It is conventional to call such programs ``daemons''\cite{Raymond_1996}.
The middle computer is connected two two additional peripherals, and runs daemons for each.
Besides communicating with the hardware peripheral, each daemon can communicate with other programs, ``clients'',  through the network.
The four client programs shown in Figure \ref{fig:network} can each communicate with all six hardware peripherals shown.
As an example, a client running on the bottom computer could communicate with the RS232 peripheral shown in green via the following path: client $\leftrightarrow$ network switch $\leftrightarrow$ top computer $\leftrightarrow$ daemon $\leftrightarrow$ hardware peripheral.
This powerful architecture can scale down to a single computer or up to many networked computers, including fully remote operator interfaces.
This client-server architecture offers similar network capabilities to EPICS and TANGO.
As we will show, usage of standards and the creation of tooling makes this architecture accessible to instrument builders outside of large facilities.

@@

The first hardware interface barrier: Multiple incompatible interfaces are used to communicate with each component of the system.
\yaq{} provides a unified TCP/IP interface to all hardware peripherals based on the well-described Avro RPC protocol.
In particular, the trait system was introduced in pursuit of our primary goal of easing the client development process.
Clients can trust that protocols that implement a given trait will behave in similar ways.

@@

The third hardware interface barrier: Some hardware interfaces are blocking.
Monolithic orchestration software often necessitates separate threads for each component hardware interface, a fragile pattern in which small mistakes become difficult to find and critical errors.
In \yaq{}, each daemon is only responsible for a single hardware interface.
%Each \yaq{} daemon runs as a separate process (possibly not even on the same machine) and thus many of the hard concurrency problems with communicating with hardware are mitigated.
It is expected that each message call over the \yaq{} interface will return rapidly, ensuring that client applications are not blocked for extended periods of time.
This principle applies even when the message starts an action that might take several seconds to complete, such as homing a motor or initiating a measurement for a sensor.
In these instances, the initial message simply starts the action and returns, with a separate message provided to retrieve results when they exist.
In order to know how long to wait, the ``is-daemon`` trait provides a message called ``is\_busy'' which should return "true" while the long running action is not complete, and "false" once it is finished.
Additionally, multiple clients can communicate with the same daemon simultaneously.
A complex instrument may involve multiple operators watching sensor data in real time, while one program is orchestrating the hardware and recording the data.

@@

At the top, we show several features associated with the integration between \yaq{} and the Bluesky project\cite{AllanDanielB2019a}.
Bluesky provides a powerful orchestration layer for conducting and recording data for a wide variety of experimental procedures.
Bluesky is a framework which allows users to implement their own experimental procedures.
Shown in Figure \ref{fig:foundation}, we see one such GUI for parameterizing a Bluesky procedure and a representation of the resultant data.
\texttt{yaqc-bluesky}\cite{yaqc-bluesky} provides a bridge to the Bluesky ecosystem.
Similar translation layers could be built for a variety of orchestration layers such as PyMoDAQ\cite{WeberSebastien2021a}, Instrumental\cite{Bogdanowicz2022}, or TRSpectrometer\cite{trspectrometer}.

@@

The Wright Group at UW-Madison needs to orchestrate a large variety of hardware in multidimensional scans for their complex spectroscopy experiments \cite{MukamelShaul2000a, WrightJohnCurtis2011a}.
This need for exquisite hardware control has resulted in several prior attempts at ``home-built'' orchestration software \cite{CarlsonRogerJohn1988a, MeyerKentAlbert2004b, KainSchuyler2017a, ThompsonBlaiseJonathan2018a}.
Now, using \yaq{}, the Wright Group has been able to move to Bluesky rather than inventing their own sophisticated control software ``from scratch''.
The Wright Group uses simulated hardware to enable client development away from the active laboratory computers.
The same client is used on four laser systems, each with different complements of hardware.
Moving forward, the Wright Group hopes to spend less energy developing control software and more energy developing creative spectroscopy experiments.

@@

In a \yaq{} context, new hardware can be incorporated into an instrument through the addition of a new daemon.
The \yaq{} architecture automatically simplifies hardware interface development in several ways.
First, because daemons are separate and portable programs, the development effort can be spread across the community of \yaq{} users.
Often, researchers can download an existing daemon rather than writing a new one.
Second, \yaq{} daemon development can be performed separated from the particulars of any individual client.
Often, this means that initial hardware enablement work can be done on a researchers personal machine before the new hardware peripheral is installed in the instrument.
Third, when developing trait-compliant protocols, it becomes easy to design and fully test your hardware interface.
In this instance, you even can use \texttt{yaqc-qtpy} to provide a graphical program to interact with your hardware immediately.
Finally, as discussed in Section II, \yaq{} gives you options to design using remote hardware or unusual interfaces when necessary.

@@

Figure \ref{fig:histogram} shows the distribution of unique lines of Python code written to implement each of the daemons in our current ecosystem.
While lines of code is an imperfect metric, we use it here to try and represent the amount of work required to create a daemon for a new hardware peripheral.
Most interfaces, such as Brooks MFC\cite{yaqd-brooks-mfc}, have been implemented in as less than 100 lines of Python code.
Even the most complicated daemons are implemented in about 300 lines.
Implementing \yaq{} daemons using Python is enabled through our own tooling mentioned above and the large and growing ecosystem of hardware interface tooling libraries that Python now provides\cite{}.
In our experience, the process of creating a new daemon involves about a day of work after communication with the hardware using Python is well understood.
There are currently 70 daemons in the yaq project supporting at least 50 kinds of hardware, noting that some daemons support the same hardware and others are software only.
Because \yaq{} is standards based, anyone can design and publish new daemons extending our hardware support.
A living list of all daemons and supported hardware can be found on the \yaq{} website.\cite{}

The Stahl Group at UW-Madison created a custom reactor which monitors gasses being produced or consumed in the reaction head-space.  \cite{SalazarChaseA2021a}
This reactor incorporates a collection of sensitive pressure transducers and a single heating process value under computer control.
\yaq{} daemons are used to interface with each sensor and the heater controller.
Recently, experimentalists have been attempting reactions involving smaller, slower, pressure changes.
A fundamental flaw in the initial analog to digital converter board was revealed by these attempts.
As a result, a new digitizer has been purchased.
This new converter will be incorporated into the existing reactor without modifying the existing graphical user interface and data recording program, minimizing downtime.

@@

Years after the original researchers leave, large monolithic acquisition programs become unknowable, undocumented, and unmaintained.
A graduate student discovers a long standing hard-coded conversion factor that is incorrect years after implementation.
Scientists resort to sourcing a replacement for an old, broken oscilloscope due to their reliance of their software on that particular interface; newer, cheaper oscilloscopes are readily available.
A graduate student is forced to meticulously reverse engineer the LabVIEW codebase that they inherited in order to understand the details of their experiment.
Software developers refer to the extra effort required to modify or fix large unmaintained codebases as ``technical debt''.\cite{}
Technical debt grows especially fast in academic environments where graduate students are involved in projects for a limited time.

@@

In \yaq{}, each component of an instrument can be developed and distributed separately.
For example, two different instruments might happen to use the same temperature sensor.
Because the temperature sensor daemon is its own independent program, both instruments can benefit from the same daemon.
As \yaq{} grows, the "ecosystem" of existing daemons means that future instruments become easier and easier to develop.
Growing this ecosystem is a collaborative effort where many \yaq{} users create portable daemons that they need and share them for the community to use and improve.

>>>>

\documentclass[aip, amsmath, amssymb, reprint,]{revtex4-1}

@@

Modern instrumentation development often involves incorporation of many dissimilar hardware peripherals into a single unified instrument.
Increasing hardware availability has brought greater instrument complexity to small research groups.
The flexibility promised to instrument makers by large hardware availability stretches the capability of traditional monolithic orchestration software.
In many cases, a lack of software flexibility leads creative researchers to feel frustrated and unable to perform the experiments they envision.
Highly modular hardware components enable extreme experimental flexibility.
Herein we describe \yaq{}, a software project defining a new standard way of communicating with diverse hardware peripherals.
\yaq{} encourages a highly modular approach to experimental software development, which is well suited to address the experimental flexibility needs of such instruments.
\yaq{} is architected to overcome hardware communication barriers that are insurmountable with typical experimental software.
A large number of hardware peripherals are already supported with tooling available to create more.
The \yaq{} standard enables collaboration between multiple research groups, increasing code quality while lowering development effort.

@@

Instrumentation development is a key part of the scientific enterprise.
Novel instruments are typically constructed of many individual components, purchased and home-built.
Orchestration software must communicate with each hardware component in the course of a scientific experiment.
This can involve utilization of many interfaces: NI DAQmx\cite{nidaqmx}, SCPI\cite{scpi}, ModBus\cite{modbus}, PICam\cite{picam}, Thorlabs APT\cite{thorlabs_apt}, among many others.
The challenge of integrating all of these is a frustrating piece of the modern instrument development process.
Weeks can be spent just integrating one new component into an existing project.
In small academic labs, these software interfaces are typically created by student researchers without software development experience.
Student researchers rarely focus on software reusability, and a lack of maitenence and documentation can make such software more difficult to use as time goes on.
Scientists may struggle to rapidly innovate on their experimental design when each hardware addition requires major software development.

Some large user-facilities have addressed interface complexity via the adoption of unified standards, such as EPICS \cite{DalesioLR1991a} or TANGO \cite{AGotz1999TANGOA}.
The unified standards define a network interface for any hardware component.
Orchestration software can target these unified standards for reading and writing hardware state.
Small background services are written to translate the myriad component interfaces into the standard EPICS IOCs and TANGO Devices.
These programs are performant, open source, and have huge libraries of exisiting hardware interface support, but require expert management to set up and provide descriptions via a separate server program.
In our experience EPICS and TANGO do not scale well to single-investigator lab environments.

@@

Here, we discuss the design of the \yaq{} standard in the context of four challenges facing instrument designers.
First, we discuss how particularly challenging hardware interfaces can become seemingly insurmountable barriers to software control.
Next, we discuss how inflexible orchestration software can limit experimental creativity.
Next, we focus on challenges that arise when enhancing or modifying existing instruments with new hardware.
Finally, we discuss the heavy software maintenance burden that many instrument designers face.
In each case, we will highlight how the \yaq{} project is designed to alleviate that challenge.
Several case studies provide a view into the flexible ways that \yaq{} can be applied to perform different scientific experiments.
This paper provides an overview of the concepts and motivations behind \yaq{}.
Refer to the \yaq{} website for a formal specification of the \yaq{} standard.\cite{yaq_fyi}

@@

In this section we describe the architecture of the \yaq{} framework in light of three major barriers that we have encountered in scientific instrumentation development.
First barrier: Multiple interfaces are used to communicate with each component of the system.
A fully automated system must be able to use all of these interfaces, a daunting task for scientists who do not specialize in software development.
Second barrier: Certain specialty hardware have inconvenient interface requirements.
A camera will only work with an obsolete interface card and drivers for Windows XP.
A data acquisition manufacturer provides an Application Programmer Interface (API) that only works in Python 3.7.
A graduate student wishes to drive several stepper motors using a Raspberry Pi.
Third barrier: Some hardware interfaces are blocking.
A graphical user interface stalls while waiting for a camera to collect data.
Custom orchestration software needs to be closed before the manufacturers configuration software can be used.
A graduate student finds themself needing to master advanced concepts in concurrency in order to orchestrate many motors performantly.

@@

Figure \ref{fig:network} diagrams the \yaq{} architecture.
Here, we show three different computers connected via an ethernet network.
The top and bottom computer are connected to monitors for interactive use while the middle computer is only accessible via the network.
This might represent a complex scientific instrument involving several operator terminals as well as embedded computers.
At the top, a single computer is connected to four hardware peripherals through RS232, TTL, USB, and PCI as indicated by the colored lines.
That same computer is running four separate programs, one for each peripheral.
These small, targeted, programs are managed by the operating system and run in the background.
It is conventional to call such programs ``daemons''\cite{Raymond_1996}.
The middle computer is connected two two additional peripherals, and runs daemons for each.
Besides communicating with the hardware peripheral, each daemon can communicate with other programs, ``clients'',  through the network.
The four client programs shown in Figure \ref{fig:network} can each communicate with all six hardware peripherals shown.
As an example, a client running on the bottom computer could communicate with the RS232 peripheral shown in green via the following path: client $\leftrightarrow$ network switch $\leftrightarrow$ top computer $\leftrightarrow$ daemon $\leftrightarrow$ hardware peripheral.
This powerful architecture can be used on a single computer or used across many networked computers, including fully remote operator interfaces.
This client-server architecture offers similar network capabilities to EPICS and TANGO.
As we will show, usage of standards and the creation of tooling makes this architecture accessible to instrument builders outside of large facilities.

@@

The first hardware interface barrier: Multiple incompatible interfaces are used to communicate with each component of the system.
\yaq{} provides a unified TCP/IP interface to all hardware peripherals based on the well-described Avro RPC protocol.
The trait system was introduced in pursuit of our primary goal of easing the client development process.
Clients can trust that protocols that implement a given trait will behave in similar ways.
The standardized \yaq{} interface hides all interface complexity away from researchers and client-side scientific code.

@@

The third hardware interface barrier: Some hardware interfaces are blocking.
Monolithic orchestration software often necessitates separate threads for each component hardware interface, a fragile pattern in which small mistakes become both critical and elusive errors.
In \yaq{}, each daemon is only responsible for a single hardware interface.
%Each \yaq{} daemon runs as a separate process (possibly not even on the same machine) and thus many of the hard concurrency problems with communicating with hardware are mitigated.
It is expected that each message call over the \yaq{} interface will return rapidly, ensuring that client applications are not blocked for extended periods of time.
This principle applies even when the message starts an action that might take several seconds to complete, such as homing a motor or initiating a measurement for a sensor.
In these instances, the initial message simply starts the action and returns, with a separate message provided to retrieve results when they exist.
In order to know how long to wait, the ``is-daemon`` trait provides a message called ``is\_busy'' which should return "true" while the long running action is not complete, and "false" once it is finished.
Additionally, multiple clients can communicate with the same daemon simultaneously.
A complex instrument may involve multiple operators watching sensor data in real time, while one program is orchestrating the hardware and recording the data.

@@

At the top, we show several features associated with the integration between \yaq{} and the Bluesky project\cite{AllanDanielB2019a}.
Bluesky provides a powerful orchestration layer for conducting and recording data for a wide variety of experimental procedures.
Shown in Figure \ref{fig:foundation}, we see one such GUI for parameterizing a Bluesky procedure and a representation of the resultant data.
\texttt{yaqc-bluesky}\cite{yaqc-bluesky} provides a bridge to the Bluesky ecosystem.
Similar translation layers could be built for a variety of orchestration layers such as PyMoDAQ\cite{WeberSebastien2021a}, Instrumental\cite{Bogdanowicz2022}, or TRSpectrometer\cite{trspectrometer}.

@@

The Wright Group at UW-Madison needs to orchestrate a large variety of hardware in multidimensional scans for their complex spectroscopy experiments \cite{MukamelShaul2000a, WrightJohnCurtis2011a}.
This need for exquisite hardware control has resulted in several prior attempts at ``home-built'' orchestration software \cite{CarlsonRogerJohn1988a, MeyerKentAlbert2004b, KainSchuyler2017a, ThompsonBlaiseJonathan2018a}.
Now, using \yaq{}, the Wright Group has been able to move to Bluesky rather than inventing their own sophisticated control software ``from scratch''.
The Wright Group uses simulated hardware to enable client development away from the active laboratory computers.
Remote client development on viritual hardware has allowed client design to proceed without delaying ongling experiments.
Clients developed by the Wright Group have proved flexible enough to be used on four laser systems, each with different complements of hardware.
Moving forward, the Wright Group will spend less energy developing control software and more energy developing creative spectroscopy experiments.

@@

In a \yaq{} context, new hardware can be incorporated into an instrument through the addition of a new daemon.
The \yaq{} architecture simplifies hardware interface development in several ways.
First, because daemons are separate and portable programs, the development effort can be spread across the community of \yaq{} users.
Often, researchers can download an existing daemon rather than writing a new one.
Second, \yaq{} daemon development can be performed separated from the particulars of any individual client.
Often, this means that initial hardware enablement work can be done on a researchers personal machine before the new hardware peripheral is installed in the instrument.
Third, when developing trait-compliant protocols, it becomes easy to design and fully test your hardware interface.
Traits are unambiguous and well-described, making an obvious target for development.
Tooling exists to verify full trait compliance, for example you can use \texttt{yaqc-qtpy} to provide a graphical program to interact with your hardware immediately.
Finally, as discussed in Section II, \yaq{} gives you options to design using remote hardware or unusual interfaces when necessary.

@@

Figure \ref{fig:histogram} shows the distribution of unique lines of Python code written to implement each of the daemons in our current ecosystem.
While lines of code is an imperfect metric, we use it here to represent the amount of work required to create a daemon for a new hardware peripheral.
Most interfaces, such as Brooks MFC\cite{yaqd-brooks-mfc}, have been implemented in as less than 100 lines of Python code.
Even the most complicated daemons are implemented in about 300 lines.
Implementing \yaq{} daemons using Python is enabled through our own tooling mentioned above and the large and growing ecosystem of hardware interface tooling libraries that Python now provides\cite{}.
In our experience, the process of creating a new daemon involves about a day of work after communication with the hardware using Python is well understood.
There are currently 70 daemons in the yaq project supporting at least 50 kinds of hardware, noting that some daemons support the same hardware and others are software only.
Because \yaq{} is standards based, anyone can design and publish new daemons extending our hardware support.
A living list of all daemons and supported hardware can be found on the \yaq{} website.\cite{}

The Stahl Group at UW-Madison created a custom reactor which monitors gasses being produced or consumed in the reaction head-space.  \cite{SalazarChaseA2021a}
This reactor incorporates a collection of sensitive pressure transducers and a single heating process value under computer control.
\yaq{} daemons are used to interface with each sensor and the heater controller.
Recently, experimentalists have been attempting reactions involving smaller, slower, pressure changes.
A fundamental flaw in the initial analog to digital converter board was revealed by these attempts.
As a result, a new digitizer has been purchased.
This new digitizer will be incorporated into the existing reactor without modifying the existing graphical user interface and data recording program, minimizing downtime.

@@

Years after the original researchers leave, large monolithic acquisition programs become unknowable, undocumented, and unmaintained.
A graduate student discovers a long standing hard-coded conversion factor that is incorrect years after implementation.
Scientists resort to sourcing a replacement for an old, broken oscilloscope due to their reliance of their software on that particular interface while newer, cheaper oscilloscopes are readily available.
A graduate student is forced to meticulously reverse engineer the LabVIEW codebase that they inherited in order to understand the details of their experiment.
Software developers refer to the extra effort required to modify or fix large unmaintained codebases as ``technical debt''.\cite{}
Technical debt grows especially fast in academic environments where graduate students are involved in projects for a limited time.

@@

In \yaq{}, each component of an instrument can be developed and distributed separately.
For example, two different instruments might happen to use the same temperature sensor.
Because the temperature sensor daemon is its own independent program, both instruments can benefit from the same daemon.
The growing ``ecosystem'' of \yaq{} daemons make future instruments easier and easier to develop.
Growing this ecosystem is a collaborative effort where many \yaq{} users create portable daemons that they need and share them with the community where they will be used and improved.
